Fetching and pre-processing landcover data

### 2015-06-26 Rmd/fetch-land-cover.Rmd
Created markdown file to select magisterial districts for agent-based model analysis.  Edited formats in older fetch-landcover.Rmd file.

# projections, etc
prjsrid <- 97490  # Albers
prj_sql <- paste("select proj4text from spatial_ref_sys where srid=", prjsrid, sep = "")
prjstr <- dbGetQuery(con, prj_sql)$proj4text
gcsid <- 4326  # GCS
gcs_sql <- paste("select proj4text from spatial_ref_sys where srid=", gcsid, sep = "")
gcsstr <- dbGetQuery(con, gcs_sql)$proj4text

@@

# Create 1 km grid from sa1kilo to rectify grids against that 
if(!file.exists(full_path(p_data, "sagrid.tif"))) {  # Wrap in if statement in case grid already made
  sql <- paste("SELECT gid, id, ST_AsText(ST_Centroid(geom)) as center FROM sa1kilo")
  geom.tab <- dbGetQuery(con, sql)
  coord.mat <- do.call(rbind, lapply(1:nrow(geom.tab), function(y) {
    strip <- gsub("POINT|\\(|\\)", "", geom.tab[y, 3])
    coord.mat <- do.call(rbind, strsplit(strsplit(strip, ",")[[1]], " "))
  }))
  class(coord.mat) <- "numeric"
  point.tab <- cbind(geom.tab[, 1:2], coord.mat)
  colnames(point.tab)[3:4] <- c("x", "y") 
  pointsXYZ <- point.tab[, c(3:4, 2)]
  sa_r <- rasterFromXYZ(pointsXYZ)
  projection(sa_r) <- sa.buf@proj4string
  sa_r <- writeRaster(sa_r, filename = full_path(p_data, "sagrid.tif"), overwrite = TRUE)
} else {
  sa_r <- raster(full_path(p_data, "sagrid.tif"))
}
```

#### Bring in landcover datasets
Let's start with the improved/fusion datasets, GLC share and geo-wiki's
```{r, eval = FALSE}
if(!file.exists(full_path(p_data, "glcsa_masked.tif"))) {  # see if file exists, to avoid redoing step
  path <- "/u/lestes/spatial_data/glc_share/"
  url <- "http://www.fao.org/geonetwork/srv/en/resources.get?id=47948&fname=GlcShare_v10_02.zip&acces=private"
  download.file(url, method = "auto", destfile = "glc_share.zip")
  unzip("glc_share.zip", exdir = "glc_share")
  glc <- raster(full_path(path, "glc_shv10_02.Tif"))
  projection(glc) <- gcsstr
  glcSA <- crop(glc, y=sa_buf_gcs, file = full_path(path, "glcSA.tif"), overwrite = TRUE)  # crop to SA
  gdalwarp(srcfile = glcSA@file@name, dstfile = full_path(path, "glcSA_alb.tif"), t_srs = prjstr, 
           tr = c(1000, 1000), r = "bilinear")
  glcSA_alb <- raster(full_path(path, "glcSA_alb.tif"))

@@

if(!file.exists(full_path(p_data, "geowikisa_masked.tif"))) {  # see if file exists, to avoid redoing step 
   # first downloaded manually from password protected geo-wiki site
  path <- "/u/lestes/spatial_data/geowiki/"
  unzip("cropland_hybrid_14052014v8.zip", exdir= "geowiki") 
  geowiki <- raster(full_path(path, "Hybrid_14052014V8.img"))
  projection(geowiki) <- gcsstr
  geowikiSA <- crop(geowiki, y=sa_buf_gcs, file = full_path(path, "geowikiSA.tif"))  # crop to extent of SA
  gdalwarp(srcfile = geowikiSA@file@name, dstfile = full_path(path, "geowikiSA_alb.tif"), t_srs = prjstr, 
           tr = c(1000, 1000), r = "bilinear")
  geowikisa_alb <- raster(full_path(path, "geowikiSA_alb.tif"))
  geowikisa <- resample(geowikisa_alb, sa_r, method = "bilinear", 
                        filename = full_path(path, "geowikisa_alb_rect.tif"))    # Warp to SA grid
  geowikisa_m <- mask(glcsa, sa_r, file = full_path(p_data, "geowikisa_masked.tif"), overwrite = TRUE)
} else {
  geowikisa_m <- raster(full_path(p_data, "geowikisa_masked.tif"))
}
```

@@

# MODIS landcover 
moddir <- "/u/lestes/spatial_data/MCD12Q1/"
if(!file.exists(full_path(moddir, "modis_type_1_alb.tif"))) {
  tiles <- expand.grid("h" = c(19, 20), "v" = c(11, 12))
  tiles <- apply(tiles, 1, function(x) paste("h", x[1], "v", x[2], sep = ""))
  url <- "http://e4ftl01.cr.usgs.gov/MOTA/MCD12Q1.051/2011.01.01/"
  tile.names <- getURL(url, verbose=TRUE, dirlistonly = TRUE) 
  tile.names1 <- strsplit(tile.names, "alt")[[1]]
  tile.names2 <- unique(substr(tile.names1, 20, 64))
  modnames <- tile.names2[sapply(tiles, function(x) grep(x, tile.names2))]
  lapply(modnames, function(x) {
    url <- paste("http://e4ftl01.cr.usgs.gov/MOTA/MCD12Q1.051/2011.01.01/", x, sep = "")
    download.file(url, method="auto", destfile = x)
  })
  dir.create("MCD12Q1")
  file.copy(from = dir(pattern = "hdf"), to = paste(getwd(), "MCD12Q1", dir(pattern = "hdf"), sep = "/"))
  file.remove(dir(pattern = "hdf"))

@@

  # Translate to geotiff
  lapply(1:5, function(x) {
    batch_gdal_translate(infiles = modis, outdir = moddir, 
                         outsuffix = paste("_LC_", x, ".tif", sep = ""), sd_index = x)  
  })
  sdnames <- lapply(1:5, function(x) dir(moddir, pattern  = paste("LC_", x, ".tif", sep = "")))

  # warp to albers, mosaic, and clip to SA extent
  lapply(1:5, function(x) {
    print(paste("processing type", x))
    gdalwarp(srcfile = sdnames[[x]], dstfile = paste("modis_type_", x, "_alb.tif", sep = ""), 
             t_srs = prjstr, tr = res(r), te = bbox(sa.buf)[1:4])
  })  
  mod_1 <- raster("modis_type1_alb.tif")
} else {
  mod_1 <- raster(full_path(moddir, "modis_type_1_alb.tif"))
}

@@

  # process masks--extract extract landcover class 2, cultivation, and class 6, plantations, resample, mask 
  # with gdal tools
  av <- c(2, 6)
  a <- "sa_lc_2009.tif"
  dang <- Sys.time() 
  salc_l <- lapply(1:2, function(j) {
    print(paste("extracting cover for", lcnm[j]))
    gdal_calc(cstr = paste("A==", av[j], sep = ""), x = list("A" = a), type = "Byte", filename = lcnm[j])
    nm <- full_path(p_data, paste(gsub("\\.tif", "", lcnm[j]), "_masked.tif", sep = ""))
    ext <- sapply(c("xmin", "ymin", "xmax", "ymax"), function(x) slot(extent(sa_r), x))
    print(paste("warping and masking", gsub(p_data, "", nm)))
    gdalwarp(srcfile = lcnm[j], t_srs = projection(sa_r), dstfile = nm, r = "average", ot = "Float32", 
             te = ext, srcnodata = 255, dstnodata = 255, tr = c(1000, 1000), of = "GTiff", verbose = TRUE,
             overwrite = TRUE)
    file.remove(lcnm[j])
    r <- raster(nm)
  })
  Sys.time() - dang

@@

  # system call to gdal_calc.py to extract landcover class 2, cultivation
  kznlconm <- "kzn_cane.tif"  # file.remove(kznlconm)
  kzninnm <- strsplit(kznlc@file@name, "/")[[1]][length(strsplit(kznlc@file@name, "/")[[1]])]
  # kzninnm <- strsplit(kznlc.ss@file@name, "/")[[1]][length(strsplit(kznlc.ss@file@name, "/")[[1]])]
  pcalc <- paste("gdal_calc.py -A ", kzninnm, " --outfile=", kznlconm, 
                 " --overwrite --calc='(A==9)|(A==10)'", sep = "")
  system.time(system(pcalc))  # 49 seconds
  kzn_cane <- raster(full_path(path, kznlconm))
  # cext <- extent(c(20000, 40000, -3220000, -3200000))  # crops to small extent for examination
  # kznlc.ss <- crop(kznlc, cext, filename = full_path(kznpath, "kznlandcover_crop.tif"), overwrite = TRUE)
  # plot(kznlc.ss)
  # scane1 <- (kznlc.ss == 9); scane2 <- (kznlc.ss == 10)
  # plot(scane1, add = TRUE, col = c("transparent", "red"))
  # plot(scane2, add = TRUE, col = c("transparent", "purple"))
  # chk <- crop(kzn_cane, cext)
  # plot(chk); plot(scane1, add = TRUE, col = c("transparent", "red"), legend = FALSE) 
  # plot(scane2, add = TRUE, col = c("transparent", "red"), legend = FALSE)
  # file.remove("kznlandcover_crop.tif")
  # rm(chk, kznlc.ss, scane1, scane2)
  file.remove("kznlandcover.tif")

@@

  # assign percentages
  minset <- c(100, 10)  # assume 10% is minimum crop cover for MODIS IGBP mosaic class
  meanset <- c(100, 35)  # assume 35% is mean crop cover
  maxset <- c(100, 60)  # assume 60% is mean crop cover  
  modsa_min <- assignLCPct(mod_1_cropb, minset, fname = full_path(p_data, "mod_1_min.tif"))
  modsa_mu <- assignLCPct(mod_1_cropb, meanset, fname = full_path(p_data, "mod_1_mu.tif"))
  modsa_max <- assignLCPct(mod_1_cropb, maxset, fname = full_path(p_data, "mod_1_max.tif"))

  # aggregate, resample, mask to SA grid
  modsa_list <- lapply(list(modsa_min, modsa_mu, modsa_max), function(x) {
    nm <- gsub("//.tif", "", x@file@name)
    print(paste("Processing", nm))
    print("...aggregating")
    agg <- aggregate(x, fact = 2, na.rm = TRUE)  
    print("...resampling")
    crop1km <- resample(agg, sa_r, method = "bilinear")
    print("...masking")
    crop1km <- mask(crop1km, mask = sa_r, filename = paste(nm, "_1kmrect.tif", sep = ""), overwrite = TRUE)
    return(crop1km)
  })

>>>>

# projections, etc
prjsrid <- 97490  # Albers
prj_sql <- paste0("select proj4text from spatial_ref_sys where srid=", prjsrid)
prjstr <- dbGetQuery(con, prj_sql)$proj4text
gcsid <- 4326  # GCS
gcs_sql <- paste0("select proj4text from spatial_ref_sys where srid=", gcsid)
gcsstr <- dbGetQuery(con, gcs_sql)$proj4text

@@

# Create 1 km grid from sa1kilo to rectify grids against that 
if(!file.exists(full_path(p_data, "sagrid.tif"))) {  # already made? 
  sql <- paste("SELECT gid, id, ST_AsText(ST_Centroid(geom)) as center", 
               "FROM sa1kilo")
  geom.tab <- dbGetQuery(con, sql)
  coord.mat <- do.call(rbind, lapply(1:nrow(geom.tab), function(y) {
    strip <- gsub("POINT|\\(|\\)", "", geom.tab[y, 3])
    coord.mat <- do.call(rbind, strsplit(strsplit(strip, ",")[[1]], " "))
  }))
  class(coord.mat) <- "numeric"
  point.tab <- cbind(geom.tab[, 1:2], coord.mat)
  colnames(point.tab)[3:4] <- c("x", "y") 
  pointsXYZ <- point.tab[, c(3:4, 2)]
  sa_r <- rasterFromXYZ(pointsXYZ)
  projection(sa_r) <- sa.buf@proj4string
  sa_r <- writeRaster(sa_r, filename = full_path(p_data, "sagrid.tif"), 
                      overwrite = TRUE)
} else {
  sa_r <- raster(full_path(p_data, "sagrid.tif"))
}
```

#### Bring in landcover datasets
Let's start with the improved/fusion datasets, GLC share and geo-wiki's
```{r, eval = FALSE}
if(!file.exists(full_path(p_data, "glcsa_masked.tif"))) {  # see if file exists,  
  # to avoid redoing step
  path <- "/u/lestes/spatial_data/glc_share/"
  url <- "http://www.fao.org/geonetwork/srv/en/resources.get?id=47948&fname=GlcShare_v10_02.zip&acces=private"
  download.file(url, method = "auto", destfile = "glc_share.zip")
  unzip("glc_share.zip", exdir = "glc_share")
  glc <- raster(full_path(path, "glc_shv10_02.Tif"))
  projection(glc) <- gcsstr
  glcSA <- crop(glc, y=sa_buf_gcs, file = full_path(path, "glcSA.tif"), 
                overwrite = TRUE)  # crop to SA
  gdalwarp(srcfile = glcSA@file@name, 
           dstfile = full_path(path, "glcSA_alb.tif"), t_srs = prjstr, 
           tr = c(1000, 1000), r = "bilinear")
  glcSA_alb <- raster(full_path(path, "glcSA_alb.tif"))

@@

if(!file.exists(full_path(p_data, "geowikisa_masked.tif"))) {  # see if file exists, to avoid redoing step 
   # first downloaded manually from password protected geo-wiki site
  path <- "/u/lestes/spatial_data/geowiki/"
  unzip("cropland_hybrid_14052014v8.zip", exdir= "geowiki") 
  geowiki <- raster(full_path(path, "Hybrid_14052014V8.img"))
  projection(geowiki) <- gcsstr
  geowikiSA <- crop(geowiki, y=sa_buf_gcs, 
                    file = full_path(path, "geowikiSA.tif"))  # crop to SA
  gdalwarp(srcfile = geowikiSA@file@name, 
           dstfile = full_path(path, "geowikiSA_alb.tif"), t_srs = prjstr, 
           tr = c(1000, 1000), r = "bilinear")
  geowikisa_alb <- raster(full_path(path, "geowikiSA_alb.tif"))
  geowikisa <- resample(geowikisa_alb, sa_r, method = "bilinear", # Warp to SA
                        filename = full_path(path, "geowikisa_alb_rect.tif"))    
  geowikisa_m <- mask(glcsa, sa_r, 
                      file = full_path(p_data, "geowikisa_masked.tif"), 
                      overwrite = TRUE)
} else {
  geowikisa_m <- raster(full_path(p_data, "geowikisa_masked.tif"))
}
```

@@

# MODIS landcover 
moddir <- "/u/lestes/spatial_data/MCD12Q1/"
if(!file.exists(full_path(moddir, "modis_type_1_alb.tif"))) {
  tiles <- expand.grid("h" = c(19, 20), "v" = c(11, 12))
  tiles <- apply(tiles, 1, function(x) paste("h", x[1], "v", x[2], sep = ""))
  url <- "http://e4ftl01.cr.usgs.gov/MOTA/MCD12Q1.051/2011.01.01/"
  tile.names <- getURL(url, verbose=TRUE, dirlistonly = TRUE) 
  tile.names1 <- strsplit(tile.names, "alt")[[1]]
  tile.names2 <- unique(substr(tile.names1, 20, 64))
  modnames <- tile.names2[sapply(tiles, function(x) grep(x, tile.names2))]
  lapply(modnames, function(x) {
    url <- paste0("http://e4ftl01.cr.usgs.gov/MOTA/MCD12Q1.051/2011.01.01/", x)
    download.file(url, method="auto", destfile = x)
  })
  dir.create("MCD12Q1")
  file.copy(from = dir(pattern = "hdf"), 
            to = paste(getwd(), "MCD12Q1", dir(pattern = "hdf"), sep = "/"))
  file.remove(dir(pattern = "hdf"))

@@

  # Translate to geotiff
  lapply(1:5, function(x) {
    batch_gdal_translate(infiles = modis, outdir = moddir, 
                         outsuffix = paste("_LC_", x, ".tif", sep = ""), 
                         sd_index = x)  
  })
  sdnames <- lapply(1:5, function(x) dir(moddir, 
                                         pattern = paste0("LC_", x, ".tif")))

  # warp to albers, mosaic, and clip to SA extent
  lapply(1:5, function(x) {
    print(paste("processing type", x))
    gdalwarp(srcfile = sdnames[[x]], 
             dstfile = paste0("modis_type_", x, "_alb.tif"), 
             t_srs = prjstr, tr = res(r), te = bbox(sa.buf)[1:4])
  })  
  mod_1 <- raster("modis_type1_alb.tif")
} else {
  mod_1 <- raster(full_path(moddir, "modis_type_1_alb.tif"))
}

@@

  # process masks--extract extract landcover class 2, cultivation, and class 6,
  # plantations, resample, mask with gdal tools
  av <- c(2, 6)
  a <- "sa_lc_2009.tif"
  dang <- Sys.time() 
  salc_l <- lapply(1:2, function(j) {
    print(paste("extracting cover for", lcnm[j]))
    gdal_calc(cstr = paste("A==", av[j], sep = ""), x = list("A" = a), 
              type = "Byte", filename = lcnm[j])
    nm <- full_path(p_data, 
                    paste(gsub("\\.tif", "", lcnm[j]), "_masked.tif", sep = ""))
    ext <- sapply(c("xmin", "ymin", "xmax", "ymax"), function(x) {
      slot(extent(sa_r), x)
    })
    print(paste("warping and masking", gsub(p_data, "", nm)))
    gdalwarp(srcfile = lcnm[j], t_srs = projection(sa_r), dstfile = nm, 
             r = "average", ot = "Float32", te = ext, srcnodata = 255, 
             dstnodata = 255, tr = c(1000, 1000), of = "GTiff", verbose = TRUE,
             overwrite = TRUE)
    file.remove(lcnm[j])
    r <- raster(nm)
  })
  Sys.time() - dang

@@

  # system call to gdal_calc.py to extract landcover class 2, cultivation
  kznlconm <- "kzn_cane.tif"  # file.remove(kznlconm)
  ss <- strsplit(kznlc@file@name, "/")
  kzninnm <- ss[[1]][length(ss[[1]])]
  # kzninnm <- strsplit(kznlc.ss@file@name, "/")[[1]][length(strsplit(kznlc.ss@file@name, "/")[[1]])]
  pcalc <- paste("gdal_calc.py -A ", kzninnm, " --outfile=", kznlconm, 
                 " --overwrite --calc='(A==9)|(A==10)'", sep = "")
  system.time(system(pcalc))  # 49 seconds
  kzn_cane <- raster(full_path(path, kznlconm))
  # cext <- extent(c(20000, 40000, -3220000, -3200000))  # crops to small extent for examination
  # kznlc.ss <- crop(kznlc, cext, filename = full_path(kznpath, "kznlandcover_crop.tif"), overwrite = TRUE)
  # plot(kznlc.ss)
  # scane1 <- (kznlc.ss == 9); scane2 <- (kznlc.ss == 10)
  # plot(scane1, add = TRUE, col = c("transparent", "red"))
  # plot(scane2, add = TRUE, col = c("transparent", "purple"))
  # chk <- crop(kzn_cane, cext)
  # plot(chk); plot(scane1, add = TRUE, col = c("transparent", "red"), legend = FALSE) 
  # plot(scane2, add = TRUE, col = c("transparent", "red"), legend = FALSE)
  # file.remove("kznlandcover_crop.tif")
  # rm(chk, kznlc.ss, scane1, scane2)
  file.remove("kznlandcover.tif")

@@

  # assign percentages
  minset <- c(100, 10)  # assume 10% is min crop cover for MODIS mosaic class
  meanset <- c(100, 35)  # assume 35% is mean crop cover
  maxset <- c(100, 60)  # assume 60% is mean crop cover  
  modsa_min <- assignLCPct(mod_1_cropb, minset, 
                           fname = full_path(p_data, "mod_1_min.tif"))
  modsa_mu <- assignLCPct(mod_1_cropb, meanset, 
                          fname = full_path(p_data, "mod_1_mu.tif"))
  modsa_max <- assignLCPct(mod_1_cropb, maxset, 
                           fname = full_path(p_data, "mod_1_max.tif"))

  # aggregate, resample, mask to SA grid
  modsa_list <- lapply(list(modsa_min, modsa_mu, modsa_max), function(x) {
    nm <- gsub("//.tif", "", x@file@name)
    print(paste("Processing", nm))
    print("...aggregating")
    agg <- aggregate(x, fact = 2, na.rm = TRUE)  
    print("...resampling")
    crop1km <- resample(agg, sa_r, method = "bilinear")
    print("...masking")
    crop1km <- mask(crop1km, mask = sa_r, filename = paste0(nm, "_1kmrect.tif"),
                    overwrite = TRUE)
    return(crop1km)
  })
